Resolvable Ambiguity

### 2019-07-10 paper/paper.tex
Fix build
Polish in Sec 4
First changes of intro

\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangebag}{\mathit{rangebag}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangeset}{\mathit{rangeset}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}
\newcommand{\regex}{\mathit{Reg}}
\newcommand{\reqpl}{\underline{(}}
\newcommand{\reqpr}{)}
\newcommand{\reqp}[1]{\reqpl#1\reqpr}
\newcommand{\pospl}{(}
\newcommand{\pospr}{)}
\newcommand{\posp}[1]{\pospl#1\pospr}
\newcommand{\dfa}{\mathit{dfa}} % Go from label to its corresponding dfa
\newcommand{\passthrough}{\mathit{passthrough}} % Go from label to a set of (label x bool), where each element denotes a child production that can be an only child (i.e., no other terminals whatsoever), and the bool states whether it requires parentheses
\newcommand{\nt}{\mathit{nt}} % Go from label to the non-terminal to which that labelled production belongs

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
% Why?

Ever since the early 60s, the problem of deciding if a context-free
grammar is unambiguous or not has been known to be undecidable. As a
consequence, a large number of restricted grammars have been developed
to guarantee that a language definition is unambiguous. This
traditional view of only allowing unambiguous grammars has until
recently been taken for granted as the only truth: \emph{the} way of
how the syntax of a language must be defined. However, up front
unambiguous grammars can force language designers to make arbitrary
choice to disambiguate the language, although the natural choice may be to
postpone the decision to the programmer. Moreover, an important
objective of domain-specific language development is to construct
languages by composing and extending existing
languages. Unfortunately, the composition of restricted unambiguous
grammars easily results in ambiguous grammars, which again are
undecidable. In this paper, we depart from the traditional view of
unambiguous grammar design, and allow ambiguities to be delayed until
compile time, allowing the user to perform the disambiguation. A
natural decision problems follows: given a language definition, can a
user always disambiguate an ambiguous program?  We introduce and
formalize this fundamental problem---called the \emph{resolvable
  ambiguity problem}---and divide it into separate static and dynamic
resolvability problems. We prove soundness and completeness for a
restricted language in the static case, and soundness in the general
case for dynamic resolvability. The approach is evaluated through
three separate case studies, covering both a large
existing programming language, and the composability of
domain-specific languages.
%A traditional view when designing programming languages is that syntax
%definitions should be unambiguous: only unambiguous grammars are used
%to construct language specific parsers that either rejects a program,
%or accepts it and produces a unique parse tree. However,
\end{abstract}

@@

When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments. However, how to make such compositions well-behaved is not always obvious. In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous. Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous programs. The former is well explored in the form of heuristics for ambiguity detection (e.g. \cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g. \cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny. This seems to stem from the view that ambiguity has no place in programming language syntax, which is highly prevalent (e.g. \cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).

Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g. \cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}. However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity. Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019} but not yet explored, is to automatically generate several unambiguous programs, one for each possible interpretation, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation. For example, in a language without precedence, \verb|'1 + 2 * 3'| is ambiguous, but the two possible interpretations can be written unambiguously as \verb|'(1 + 2) * 3'| and \verb|'1 + (2 * 3)'| respectively.

In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists. For example, in a language with semicolon for both sequential composition and element separation in a list (e.g. OCaml), ``$[1; 2]$'' is ambiguous: it is either a list of two elements, or a list of one element, namely the sequential composition of $1$ and $2$. The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative. This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar. We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot. We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}. This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.

@@

\noindent We say that the automaton accepts the word $a_0a_1\ldots a_n$ iff there is such a succesful run.

@@

\begin{description}
\item[Static resolvability.] Given an abstract parser $P$, determine whether $\rho(P)$.
\item[Dynamic resolvability.] Given a word $w \in L(P)$ for some abstract parser $P$, determine whether $\rho_P(w)$.
\end{description}

@@

% todo: improve label (where used?)
\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure}
  A decision procedure $f$ solving the static resolvability problem is sound iff

@@

In practice, we may additionally be interested in a decision procedure
$g$ such that $g(P) \implies \lnot \rho(P) \text{ for all abstract
  parsers } P$. It turns out, we can give $g$ for the parse languages
studied in following sections.

\begin{definition}[Completeness of Static Resolvability]
  A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff

@@

\noindent Similarly, for the dynamic problem:

\begin{definition}
  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the dynamic resolvability problem is:
  \begin{itemize}
  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$ and $w$.
  \end{itemize}
  \label{def:dynamic-procedure}
\end{definition}

@@

Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees. Each element corresponds to the word ``$(1 + 2) * 3$'' in $L(G'_D)$. Note that the word in $L(G_D)$ is ambiguous, and that there are other words in $L(G'_D)$ that correspond to the same element in $L(T_D)$, e.g., ``$((1 + 2)) * 3$'' and ``$(1 + 2) * (3)$''. As a memory aid, the primed versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the (likely ambiguous) straightforward translations (i.e., ignoring marks) from $D$. We will generally refer to elements of $L(G_D)$ as $w$, $L(G'_D)$ as $w'$, $L(T_D)$ as $t$, and $L(T'_D)$ as $t'$.

@@

$$
\begin{array}{rcl}
\parse(w') & = & \{ \semantic(t') \mid t' \in L(T'_D) \land \yield(t') = w' \} \\
\words(t) & = & \{ w' \mid t \in \parse(w') \} \\
\end{array}
$$

@@

We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure} and \ref{def:dynamic-procedure}. Section~\ref{sec:static} gives a partial solution to the former problem, while Section~\ref{sec:dynamic} fully solves the latter with one caveat: we only consider languages with balanced parentheses.

@@

\begin{theorem}
  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous iff\\$\forall t \in T.\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$.
\end{theorem}

@@

\noindent To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w' \in L(G'_D)$ for which $\parse(w') = \{t\}$, or prove that no such tree exists. Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exists.

@@

\begin{lemma}
  $\forall w'_1, w'_2 \in \words(t).\ \basic(w'_1) = \basic(w'_2)$, i.e., all $w'$ have the same basic word.
  \label{lemma:same-basic}
\end{lemma}
\noindent Intuitively, $\words$ first applies the ``inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word. The only terminals changed in this process are parentheses, i.e., different $w'$ can only differ in their range bags. This also implies that two trees that share a word (i.e., trees $t_1$ and $t_2$ that have a word $w'$ such that $\parse(w') \supseteq \{t_1, t_2\}$) must also share a basic word.

\begin{lemma}
  $\exists R$ such that $\exists w' \in \words(t).\ \rangebag(w') = R$ and $\forall w' \in \words(t).\\ R \subseteq \support(\rangebag(w'))$, i.e., some parentheses are required.
  \label{lemma:required-parentheses}
\end{lemma}
\noindent For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the produced parse trees. Each required range is a direct consequence of a mark on a non-terminal in $D$. We will write $w'_\bot$ for the word implied by the first quantified expression. It is unique, since its basic word is fixed by $t$, and its rangebag is a set.

\begin{lemma}
  $\exists R$ such that $\exists w' \in \words(t).\ \rangebag(w') = R$ and $\forall w' \in \words(t).\\ \support(\rangebag(w')) \subseteq R$, i.e., there is a finite set of possible parentheses.
  \label{lemma:possible-parentheses}
\end{lemma}
\noindent As mentioned in Section~\ref{sec:parse-time-disambiguation}, grouping parentheses can only be added if they exactly cover a production, which amounts to an internal node in $t$, and each tree has a finite amount of nodes. We will write $w'_\top$ for the word implied by the first quantified expression, which is also unique.

\begin{lemma}
  $\forall w'_1, w'_2 \in \words(t).\ \support(\rangebag(w'_1)) = \support(\rangebag(w'_2)) => \parse(w'_1) = \parse(w'_2)$, i.e., duplicated parentheses do not matter.
  \label{lemma:rangeset-equality}
\end{lemma}
\noindent We first note that in version 1 and 2 there are no parentheses in $D$, i.e., all parentheses in $G'_D$ are introduced as grouping parentheses. Each pair of parentheses thus corresponds to a single grouping node $g$ in a tree in $L(T'_D)$. Duplicated grouping parentheses correspond to nested grouping nodes, all of which are removed by $\semantic$, thus producing identical sets of trees in $L(T_D)$. For example, \verb|((1 + 2)) * 3| has the same interpretations as \verb|(1 + 2) * 3|.

@@

Lemma~\ref{lemma:rangeset-equality} suggests a partition of the words in $\words(t)$ for any given $t \in L(T_D)$; group words $w'$ by their \emph{rangeset}, where $\rangeset(w') = \support(\rangebag(w'))$. These partitions can be partially ordered by subset on the rangeset, resulting in a lattice of word partitions per tree. This lattice is bounded, with top and bottom elements given by lemmas~\ref{lemma:possible-parentheses} and \ref{lemma:required-parentheses} respectively. Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word. For example, Fig.~\ref{fig:lattice1} contains the lattice for the tree $a(a(n(1)+n(2))+n(3))$ (from our running example, defined in Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}). Each partition is represented by the word whose rangebag is a set. To reduce clutter, we do not draw the partitions that have grouping parentheses around the entire word. This outermost possible pair is uninteresting since it is always allowed, and would double the size of the figure if it was included.

@@

\begin{lemma}
  Given $t \in L(T_D)$, $w'_\top, w'_\bot \in \words(t)$ such that $\forall w'.\ w' \in \words(t) => \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top)$ the following holds:\\
  $\forall w'.\ \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top) => w' \in \words(t)$. \label{lemma:top-bottom-determine}
\end{lemma}

\noindent Lemmas~\ref{lemma:possible-parentheses} and \ref{lemma:required-parentheses} guarantee the existence of two such words $w'_\top$ and $w'_\bot$.

Finally, we show the difference between versions~1 and 2 in the lattice setting: in version~1, the rangeset of the bottom word is the empty set:

\begin{lemma}
  In version~1, $rangeset(w'_\bot) = \emptyset$ for all trees, where $w'_\bot$ is given by Lemma~\ref{lemma:required-parentheses}. \label{lemma:version1-bot}
\end{lemma}

@@

\begin{theorem}
  In versions~1 and 2, given a $t \in L(T_D)$ with a lattice determined by $w_\bot$ and $w_\top$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w_\bot) \subseteq \rangeset(w'_\bot) \land \rangeset(w'_\top) \subseteq \rangeset(w_\top) => \neg \exists w.\ \parse(w) = \{t\}$. \label{lemma:lattice-containment-implies-unresolvable}
\end{theorem}

@@

\begin{theorem}
  In version 1, given a $t \in L(T_D)$ with a lattice determined by $w_\bot$ and $w_\top$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w'_\bot) \subseteq \rangeset(w_\bot) \land \rangeset(w_\top) \subseteq \rangeset(w'_\top) <=> \neg \exists w.\ \parse(w) = \{t\}$.
\end{theorem}

\begin{proof}
  The left-to-right implication is already given in Theorem~\ref{lemma:lattice-containment-implies-unresolvable}, and Lemma~\ref{lemma:version1-bot} together with $\basic(w_\bot) = \basic(w'_\bot)$ implies that $w_\bot = w'_\bot$. We thus need to show that $\neg \exists w.\ \parse(w) = \{t\} => \exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w'_\top) \subseteq \rangeset(w_\top)$.

  If all words in $\words(t)$ are ambiguous, then $\exists t' \in L(T_D).\ t' \neq t \land t' \in \parse(w_\top)$. Lemma~\ref{lemma:same-basic} gives $\basic(w_\bot) = \basic(w'_\bot)$. The definition of $w'_\top$ (Lemma~\ref{lemma:possible-parentheses}) gives $\rangeset(w_\top) \subseteq \rangebag(w'_\top)$, but since the rangebag of $w'_\top$ is a set, this also implies that $\rangeset(w_\top) \subseteq \rangeset(w'_\top)$.
\end{proof}

@@

Conceptually, each non-terminal represents a choice of which child production should replace it, and whether that production requires parentheses around it. We now make this explicit, to ensure that standard operations on finite automata produce the correct result. We replace each $N_m$ with a sum of the labels of productions in $N$, where each label has a subscript $\top$ or $\bot$ if parentheses are required or possible, respectively. For example, $E_{\{s\}}$ is replaced with $s_\top + v_\bot$.

@@

\noindent Similarly, for every edge $p \xrightarrow{l_\bot} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\pospl', +(p, q)} p'$ and $q' \xrightarrow{'\pospr', -(p, q)} q$.

>>>>

\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangebag}{\mathit{rangebag}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangeset}{\mathit{rangeset}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\topw}{\top_w} % go from a tree to its top word
\newcommand{\botw}{\bot_w} % go from a tree to its bottom word
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}
\newcommand{\regex}{\mathit{Reg}}
\newcommand{\reqpl}{\underline{(}}
\newcommand{\reqpr}{)}
\newcommand{\reqp}[1]{\reqpl#1\reqpr}
\newcommand{\pospl}{(}
\newcommand{\pospr}{)}
\newcommand{\posp}[1]{\pospl#1\pospr}
\newcommand{\dfa}{\mathit{dfa}} % Go from label to its corresponding dfa
\newcommand{\passthrough}{\mathit{passthrough}} % Go from label to a set of (label x bool), where each element denotes a child production that can be an only child (i.e., no other terminals whatsoever), and the bool states whether it requires parentheses
\newcommand{\labelnt}{\mathit{nt}} % Go from label to the non-terminal to which that labelled production belongs
\newcommand{\labelregex}{\mathit{regex}} % Go from label the regex of the production with that label

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
A common point of view when designing grammars for parsing is that
the grammar definition has to be unambiguous.  However, up front unambiguous grammars
can force language designers to make arbitrary choice to disambiguate
the language. In this paper, we depart from the traditional view of
unambiguous grammar design, and allow ambiguities to be delayed until
compile time, allowing the user to perform the disambiguation. A
natural decision problem follows: given a language definition, can a
user always disambiguate an ambiguous program?  We introduce and
formalize this fundamental problem---called the \emph{resolvable
  ambiguity problem}---and divide it into separate static and dynamic
resolvability problems. We prove soundness and completeness for a
restricted language in the static case, and soundness in the general
case for dynamic resolvability. The approach is evaluated through
two separate case studies, covering both a large existing
programming language, and the composability of domain-specific
languages.
\end{abstract}

@@

% Motivate why unambiguous parsing is a good (and important) idea.
Ever since the early 60s, the problem of deciding if a context-free
grammar is ambiguous has been known to be
undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
a large number of restricted grammars have been developed to guarantee
that a language definition is unambiguous. This traditional view of
only allowing unambiguous grammars has been taken for granted as the
only truth: \emph{the} way of how the syntax of a language must be
defined
%This seems to stem
%from the view that ambiguity has no place in programming language
%syntax, which is highly prevalent
~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
%
However, recent work by
~\citet{palmkvistCreatingDomainSpecificLanguages2019} shows why
carefully designed ambiguous syntax definitions are in some cases
preferable compared to completely unambiguous
definitions. Specifically, instead of forcing a language designers to
make arbitrary choice to disambiguate the language, a more natural
choice may be to postpone the decision to the programmer.

% Motivation II, compising DSLs
Another area---where ambiguous grammars naturally arises---is
domain-specific language development.  An important
objective of domain-specific language development is to construct
languages by composing and extending existing
languages.
Unfortunately, the composition of restricted
%grammars easily results in ambiguous grammars, which again are
%undecidable.
%
%constructing domain-specific programming languages there is often
%a desire to reuse work, to compose smaller language
%fragments.
%However, how to make such compositions well-behaved is not
%always obvious.
%In particular, when constructing new domain-specific languages from
%existing languages, by composing multiple syntax definitions,
grammars easily produces
an ambiguous grammar, even if all of the original grammars are
unambiguous. Approaches for solving this problem can be broadly split
into two categories: those that handle ambiguities on the
grammar-level (detection, prevention, etc.), and those that work on
particular examples of ambiguous programs. The former is well explored
in the form of heuristics for ambiguity detection
(e.g. \cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007})
or restrictions on the grammars to be composed
(e.g. \cite{kaminskiModularWellDefinednessAnalysis2013}), while the
latter has received relatively little attention.

%Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g. \cite{earleyEfficientCo%ntextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}. However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity. Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019} but not yet explored, is to automatically generate several unambiguous programs, one for each possible interpretation, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation. For example, in a language without precedence, \verb|'1 + 2 * 3'| is ambiguous, but the two possible interpretations can be written unambiguously as \verb|'(1 + 2) * 3'| and \verb|'1 + (2 * 3)'| respectively.

%In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists. For example, in a language with semicolon for both sequential composition and element separation in a list (e.g. OCaml), ``$[1; 2]$'' is ambiguous: it is either a list of two elements, or a list of one element, namely the sequential composition of $1$ and $2$. The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative.

% The problem

This paper focuses on the problems that arises if a grammar definition
is not guaranteed to be unambiguous. Specifically, if parsing a
program results in several parse trees, the grammar was obviously
ambiguous. The programmer then needs to disambiguate the program by,
for instance, inserting extra parenthesis. However, is it always
possible to resolve an ambiguous input program by augmenting the
program? We say that a program has a \emph{resolvable ambiguity} if
the ambiguity can be solved by augmenting the program. The \emph{resolvable ambiguity problem} can then be divided into several decision problems: (i) the dynamic decidability problem: given a program (or a

dynamic resolvable ambiguity problem of deciding if a context-free grammar is

This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar. We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot. We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}. This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.

@@

\noindent We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a succesful run.

@@

\begin{description}
\item[Static resolvability.] Given an abstract parser $P$, determine whether $\rho(P)$.
\item[Dynamic resolvability.] Given an abstract parser $P$ and a word $w \in L(P)$, determine whether $\rho_P(w)$.
\end{description}

@@

\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure-sound}
  A decision procedure $f$ solving the static resolvability problem is sound iff

@@

In practice, we may additionally be interested in a decision procedure
$g$ such that $g(P) \implies \lnot \rho(P) \text{ for all abstract
  parsers } P$. It turns out, we can give $g$ for the parse languages
studied in the following sections.

\begin{definition}[Completeness of Static Resolvability]\label{def:static-procedure-complete}
  A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff

@@

\noindent Similarly, we define soundness and completeness for the dynamic problem.

\begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure-sound}
  A decision procedure $f$ solving the dynamic resolvability problem is sound iff

  $f(\parse, w) \implies \rho_P(w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
\end{definition}

Analogous to the static case, for the parse languages studied in the
following sections we can also provide a decision procedure $g$ such
that $g(\parse, w) \implies \lnot \rho_P(w) \text{ for all } w \in
\Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$.

\begin{definition}[Completeness of Dynamic Resolvability]\label{def:dynamic-procedure-complete}
  A decision procedure $f$ solving the dynamic resolvability problem is complete iff

  $\rho_P(w) \implies f(\parse, w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
\end{definition}

@@

Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees. Each element corresponds to the word ``$(1 + 2) * 3$'' in $L(G'_D)$. Note that the word in $L(G_D)$ is ambiguous, and that there are other words in $L(G'_D)$ that correspond to the same element in $L(T_D)$, e.g., ``$((1 + 2)) * 3$'' and ``$(1 + 2) * (3)$''. As a memory aid, the primed versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the (likely ambiguous) straightforward translations (i.e., ignoring marks) from $D$.

@@

$$
\begin{array}{rcl}
\parse(w) & = & \{ \semantic(t) \mid t \in L(T'_D) \land \yield(t) = w \} \\
\words(t) & = & \{ w \mid t \in \parse(w) \} \\
\end{array}
$$

@@

We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure-sound}, \ref{def:static-procedure-complete}, \ref{def:dynamic-procedure-sound}, and \ref{def:dynamic-procedure-complete}. Section~\ref{sec:static} gives a partial solution to the static problem, while Section~\ref{sec:dynamic} fully solves the dynamic problem with one caveat: we only consider languages with balanced parentheses.

@@

\begin{theorem}
  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous iff\\$\forall t \in T.\ \exists w \in \T^{*}.\ \parse(w) = \{t\}$.
\end{theorem}

@@

\noindent To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w \in L(G'_D)$ for which $\parse(w) = \{t\}$, or prove that no such tree exists. Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exists.

@@

\begin{lemma}
  $\forall w_1, w_2 \in \words(t).\ \basic(w_1) = \basic(w_2)$.
  \label{lemma:same-basic}
\end{lemma}
\noindent More informally, all $w \in \words(t)$ have the same basic word. Intuitively, $\words$ first applies the ``inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word. The only terminals changed in this process are parentheses, i.e., different $w$ can only differ in their range bags. This also implies that two trees that share a word (i.e., trees $t_1$ and $t_2$ that have a word $w$ such that $\parse(w) \supseteq \{t_1, t_2\}$) must also share a basic word.

\begin{lemma}
  There exists a set $R$ such that $\exists w \in \words(t).\ \rangebag(w) = R$ and $\forall w \in \words(t).\\ R \subseteq \support(\rangebag(w))$
  \label{lemma:required-parentheses}
\end{lemma}
\noindent Recall that $\support(B)$ is the support of the bag $B$, i.e., the underlying set of elements in $B$. More informally, some parentheses are required. For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the produced parse trees. Each required range is a direct consequence of a mark on a non-terminal in $D$. We will write $\botw(t)$ for the word implied by the first quantified expression. It is unique, since its basic word is fixed by $t$, and its rangebag is a set.

\begin{lemma}
  There exists a set $R$ such that $\exists w \in \words(t).\ \rangebag(w) = R$ and $\forall w \in \words(t).\\ \support(\rangebag(w)) \subseteq R$.
  \label{lemma:possible-parentheses}
\end{lemma}
\noindent More informally, there is a finite set of possible parentheses. As mentioned in Section~\ref{sec:parse-time-disambiguation}, grouping parentheses can only be added if they exactly cover a production, which amounts to an internal node in $t$, and each tree has a finite amount of nodes. We will write $\topw(t)$ for the word implied by the first quantified expression, which is also unique.

\begin{lemma}
  $\forall w_1, w_2 \in \words(t).\ \support(\rangebag(w_1)) = \support(\rangebag(w_2)) => \parse(w_1) = \parse(w_2)$.
  \label{lemma:rangeset-equality}
\end{lemma}
\noindent More informally, duplicated parentheses do not matter. We first note that in version 1 and 2 there are no parentheses in $D$, i.e., all parentheses in $G'_D$ are introduced as grouping parentheses. Each pair of parentheses thus corresponds to a single grouping node $g$ in a tree in $L(T'_D)$. Duplicated grouping parentheses correspond to nested grouping nodes, all of which are removed by $\semantic$, thus producing identical sets of trees in $L(T_D)$. For example, \verb|((1 + 2)) * 3| has the same interpretations as \verb|(1 + 2) * 3|.

@@

Lemma~\ref{lemma:rangeset-equality} suggests a partition of the words in $\words(t)$ for any given $t \in L(T_D)$; group words $w$ by their \emph{rangeset}, where $\rangeset(w) = \support(\rangebag(w))$. These partitions can be partially ordered by subset on the rangeset, resulting in a lattice of word partitions per tree. This lattice is bounded, with top and bottom elements $\topw(t)$ and $\botw(t)$. Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word. For example, Fig.~\ref{fig:lattice1} contains the lattice for the tree $a(a(n(1)+n(2))+n(3))$ (from our running example, defined in Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}). Each partition is represented by the word whose rangebag is a set. To reduce clutter, we do not draw the partitions that have grouping parentheses around the entire word. This outermost possible pair is uninteresting since it is always allowed, and would double the size of the figure if it was included.

@@

\begin{lemma}
  Given $t \in L(T_D)$, we have that:\\
  $\forall w.\ \rangebag(\botw(t)) \subseteq \rangeset(w) \subseteq \rangebag(\topw(t)) => w \in \words(t)$. \label{lemma:top-bottom-determine}
\end{lemma}

\noindent Recall that $\rangebag = \rangeset$ for $\botw(t)$ and $\topw(t)$. Finally, we show the difference between versions~1 and 2 in the lattice setting: in version~1, the rangeset of the bottom word is the empty set:

\begin{lemma}
  In version~1, $\forall t \in L(T_D).\ \rangeset(\botw(t)) = \emptyset$. \label{lemma:version1-bot}
\end{lemma}

@@

\begin{theorem}
  In versions~1 and 2, given a $t \in L(T_D)$, $w_\bot = \botw(t)$, and $w_\top = \topw(t)$, the existence of a tree $t' \in L(T_D)$ such that $\basic(w_\bot) = \basic(\botw(t')) \land \rangeset(w_\bot) \subseteq \rangeset(\botw(t')) \land \rangeset(\topw(t')) \subseteq \rangeset(w_\top)$ implies $\neg \exists w.\ \parse(w) = \{t\}$. \label{lemma:lattice-containment-implies-unresolvable}
\end{theorem}

@@

\begin{theorem}
  In version 1, given a $t \in L(T_D)$, $w_\bot = \botw(t)$, and $w_\top = \topw(t)$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(\botw(t')) \land \rangeset(\botw(t')) \subseteq \rangeset(w_\bot) \land \rangeset(w_\top) \subseteq \rangeset(\topw(t')) <=> \neg \exists w.\ \parse(w) = \{t\}$.
\end{theorem}

\begin{proof}
  The left-to-right implication is already given in Theorem~\ref{lemma:lattice-containment-implies-unresolvable}, and Lemma~\ref{lemma:version1-bot} together with $\basic(w_\bot) = \basic(\botw(t'))$ implies that $w_\bot = \botw(t')$. We thus need to show that $\neg \exists w.\ \parse(w) = \{t\} => \exists t' \in L(T_D).\ \basic(w_\bot) = \basic(\botw(t')) \land \rangeset(\topw(t')) \subseteq \rangeset(w_\top)$.

  If all words in $\words(t)$ are ambiguous, then $\exists t' \in L(T_D).\ t' \neq t \land t' \in \parse(w_\top)$. Lemma~\ref{lemma:same-basic} gives $\basic(w_\bot) = \basic(\botw(t'))$. The definition of $\topw(t')$ (Lemma~\ref{lemma:possible-parentheses}) gives $\rangeset(w_\top) \subseteq \rangebag(\topw(t'))$, but since the rangebag of $\topw(t')$ is a set, this also implies that $\rangeset(w_\top) \subseteq \rangeset(\topw(t'))$.
\end{proof}

@@

Conceptually, each non-terminal represents a choice of which production should replace it, and whether that production requires parentheses around it. We now make this explicit, to ensure that standard operations on finite automata produce the correct result. We replace each $N_m$ with a sum of the labels of productions in $N$, where each label has a subscript $\top$ or $\bot$ if parentheses are required or possible, respectively. For example, $E_{\{s\}}$ is replaced with $s_\top + v_\bot$.

@@

\noindent Similarly, for every edge $p \xrightarrow{l_\bot} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\pospl', +(p, q)} p'$ and $q' \xrightarrow{'\pospr', -(p, q)} q$. Note the difference: for $l_\top$ we add a required parenthesis ``$\reqpl$'', while $l_\bot$ adds a possible parenthesis ``$\pospl$''.
